Open archaeology, open source?

### 2023-09-20 analysis/_03-data_methodology.qmd
Merge outline into text
Revise first results section
Trying to be a bit more focused on the results rather than questions and definitions.

---
bibliography: references.bib
---

# Methods & Materials <!-- Methods & Materials -->

@@

While our initial intention was to only list open source software, open-archaeo's scope has expanded to include all software created by and for archaeologists.
We found that many repositories do not explicitly denote an open source license, or any license at all.

```{r tbl-repo-licenses}
#| tbl-cap: Summary of licenses used by projects with GitHub repositories.
oarch |>
  hoist(gh_repo_info, license = list("license", 1, "name")) |>
  count(license) |>
  mutate(license = replace_na(license, "None detected")) |>
  arrange(-n) |>
  gt() |>
    cols_label(license = "License", n = "N")
```

>>>>

# Data and methodology

<!--
* open-archaeo
* GitHub API
* Indicate how much is on GitHub or elsewhere
  -->

@@

While our initial intention was to only list open source software, open-archaeo's scope has expanded to include all software created by and for archaeologists.

### 2023-09-20 analysis/_04-open_archaeology.qmd
Revise first results section
Trying to be a bit more focused on the results rather than questions and definitions.

# Open archaeology <!-- Results I: description of open archaeology projects -->

Our analysis documented an the widespread presence of open archaeology projects on GitHub and other online platforms.

We annotated each record with tags that describe aspects of archaeological work that each tool contributes to. We assigned tags based on how their maintainers identified their projects. The list of tags and their associated definitions is subject to grow or change as archaeologists continue to develop software for additional use cases. Right now there are 63 tags.

We also categorized records based on how each tool or resource is meant to be accessed or used. This gives us a sense of the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold. In other words, categories hint at some of the decisions that developers make regarding how they want their tools to be used and accessed.

| Category                              | Kind      | Scope                                                                                                                                                                                                        |
|:-----------|:-----------|:------------------------------------------------|
| Packages and libraries                | Software  | Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.                                                                                   |
| Standalone software                   | Software  | Software that may be operated without needing to first access an underlying platform.                                                                                                                        |
| Scripts                               | Software  | Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation. |
| Specifications, protocols and schemas | Documents | A formal data structure or framework intended to be used as a model.                                                                                                                                         |
| Products                              | Documents | Stable outcomes of creative work.                                                                                                                                                                            |
| Guides                                | Documents | An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.                                                                                          |
| Lists and datasets                    | Documents | A series of consistently organized observations assembled with purpose.                                                                                                                                      |
:Categorisation scheme used by open-archaeo {#tbl-category-defs}

We defined two kinds of categories: software and documents. Software comprise collections of instructions that dictate how a computer should operate, and documents are bounded sets of recorded information. We define three categories of software and four categories of documents. Packages and libraries are sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform. Platforms are the broader systems upon which packages or libraries are plugged into or are meant to extend upon, such as R, Python, QGIS, and Meshlab, which each have their own protocols to facilitate such integration. Standalone software are tools that users may operate without needing to first access an underlying platform. The third category of software, scripts, are sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.

The first kind of document, which we refer to as specifications, protocols and schemas, are formal data structures or frameworks intended to be used as models. Products are stable outcomes of creative work, loosely defined. Guides are educational resources or documented protocols meant to instruct readers how to apply relevant tools or techniques. And finally, lists or datasets, including open-archaeo itself, are series of consistently organized observations assembled with purpose.

## Platforms and languages

Our first set of analyses focused on these questions:

* What kinds of software are archaeologists making?
* What are the general purposes of their software?

Most projects included in open-archaeo are designed to be used as a package or library atop am existing "platform" -- for example a package for a programming language like R, or a plugin for an application like QGIS ([@fig-cats-plats]). The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes. Others create standalone software that can be run independently of such platforms, for example desktop or web apps. A significant amount of projects also comprise of datasets and non-packaged code snippets that have been made available for general use .

```{r fig-cats-plats}
#| fig-cap: Number of open archaeology projects by category and platform
#| fig-subcap: 
#|   - Categories
#|   - Platforms

oarch |>
  drop_na(category) |>
  group_by(category) |>
  summarise(n = n()) |>
  ggplot(aes(fct_reorder(category, n), n)) +
  geom_col() +
  geom_text(aes(label = category), hjust = 0, nudge_y = 1) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 300)) +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank()
  )

oarch |>
  drop_na(platform) |>
  count(platform) |>
  arrange(n) |>
  ggplot(aes(fct_reorder(platform, n), n)) +
  geom_col(colour = "#333333") +
  geom_text(aes(label = platform), hjust = 0, nudge_y = 1) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 200)) +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

R is overwhelmingly dominant as a platform, closely followed by Python [@fig-cats-plats]. Plugins for QGIS and mobile apps are also relatively popular. Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates. In the figure below, the darker grey lines are used to break up the bars by primary author, showing that many of the smaller platforms---for example Excel templates---are targeted by only one or two developers. The larger platforms tend to be more diverse.

```{r plot-platforms}
```

When we compare categories with tags, we see the general domains that each kind of resource is designed to serve. We see that packages are fairly common across the board. Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling. These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

```{r plot-tags-cats}
detail_tags <- c("Instrumental Neutron activation analysis",
                 "Harris Matrix",
                 "aDNA Simulators",
                 "Damage profilers",
                 "Metagenomic classifers",
                 "aDNA simulators",
                 "aDNA damage-aware genotypers",
                 "Low-coverage contamination estimation",
                 "Low-coverage adapted population-genetics tools",
                 "aDNA preprocessing",
                 "aDNA aligner",
                 "aDNA assembler",
                 "aDNA toolkit")

@@

## History

Archaeological software development activity has increased significantly over the years. The following charts portray the cumulative growth of code contributions committed and pushed to GitHub repositories, and the number of GitHub repositories that host archaeological software and resources.

```{r plot-gh-cum}
oarch %>% 
  mutate(
    lumped_category = recode(category,
                             "Specifications etc." = "Documents",
                             "Products" = "Documents",
                             "Guides" = "Documents",
                             "Lists and datasets" = "Documents",
                             .default = category),
    lumped_category = factor(lumped_category, c("Scripts",
                                                "Documents",
                                                "Packages and libraries",
                                                "Standalone software")) ) ->
  oarch_lumped

>>>>

---
bibliography: references.bib
---

# Open archaeology

```{=html}
<!-- Results I: description of open archaeology projects

- What kind of projects comprise open archaeology?
  x Breakdown of tags and categories
  x Breakdown of platforms and languages
  x Tags by category
  x Licenses?
- How has this changed over time?
  - Growth of open archaeology by number of commits
  - Growth of open archaeology by number of repos
  - Compare changing proportions of each category on a year by year basis
  - Identify periods associated with different rates of growth, and relate these with different general attitudes in the history of digital archaeology
-->
```
As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists, primarily software but also various forms of open document.
@tbl-categories summarises the categories included.

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects included in open-archaeo
tribble(
  ~category,                               ~kind,       ~scope,
  "Packages and libraries",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products",                              "Documents", "Stable outcomes of creative work.",
  "Guides",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(category, kind, scope) |>
  group_by(kind) |>
  arrange(-n) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase)
```

```{r num-p_platform}
p_platform <- sum(!is.na(oarch$platform)) / nrow(oarch)
```

Most projects (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language or a plugin for an application.
The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets and non-packaged code snippets that have been made available for general use.

```{r tbl-platforms}
#| tbl-cap: Platforms and programming languages used by open archaeology projects
oarch |>
  drop_na(platform) |>
  count(platform) |>
  mutate(
    significant = n > 1,
    platform = if_else(significant, platform, "Other")
  ) |>
  group_by(platform) |>
  summarise(n = sum(n)) |>
  arrange(platform == "Other", -n) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase)
```

```{r num-p_platform_r}
p_platform_r <- sum(oarch$platform == "R", na.rm = TRUE) / nrow(oarch)
```

The statistical programming language R is overwhelmingly the most common platform, representing `r percent(p_platform_r)` of projects in open-archaeo.
Python, another programming language, is also relatively popular, as are plugins for the open source geographic information system QGIS.
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

At first glance, the relative popularity of R versus Python is perhaps surprising; Python is regularly ranked as the most popular programming language in the world, with R a distant runner-up.
However, it accords with the popularity of R as a tool for data analysis in archaeology [@schmidt2020] and other scientific disciplines [@lai2019].

We also annotated each record with 'tags' that describe aspects of archaeological work that each tool contributes to [@fig-tags].
When we compare categories with tags, we see the general domains that each kind of resource is designed to serve.
We see that packages are fairly common across the board.
Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling.
These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

<!-- TODO: can we say anything more concrete about the tags? -->

```{r fig-tags}
detail_tags <- c("Instrumental Neutron activation analysis",
                 "Harris Matrix",
                 "aDNA Simulators",
                 "Damage profilers",
                 "Metagenomic classifers",
                 "aDNA simulators",
                 "aDNA damage-aware genotypers",
                 "Low-coverage contamination estimation",
                 "Low-coverage adapted population-genetics tools",
                 "aDNA preprocessing",
                 "aDNA aligner",
                 "aDNA assembler",
                 "aDNA toolkit")

@@

<!-- TODO: expand. -->
We found that many repositories do not explicitly denote an open source license, or any license at all [@tbl-repo-licenses].

```{r tbl-repo-licenses}
#| tbl-cap: Summary of licenses used by projects with GitHub repositories.
oarch |>
  hoist(gh_repo_info, license = list("license", 1, "name")) |>
  count(license) |>
  mutate(license = replace_na(license, "None detected")) |>
  arrange(-n) |>
  gt() |>
    cols_label(license = "License", n = "N")
```

Archaeological software development activity has increased significantly over the years.
@fig-github-cumulative shows the cumulative growth of code contributions committed and pushed to GitHub repositories, and the number of GitHub repositories that host archaeological software and resources.

```{r fig-github-cumulative}
oarch %>% 
  mutate(
    lumped_category = recode(category,
                             "Specifications etc." = "Documents",
                             "Products" = "Documents",
                             "Guides" = "Documents",
                             "Lists and datasets" = "Documents",
                             .default = category),
    lumped_category = factor(lumped_category, c("Scripts",
                                                "Documents",
                                                "Packages and libraries",
                                                "Standalone software")) ) ->
  oarch_lumped

### 2023-09-20 analysis/openarchaeo_collaboration.qmd
Merge bibliographies
Revise first results section
Trying to be a bit more focused on the results rather than questions and definitions.

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo (https://open-archaeo.info/), a curated list of 500+ active open projects in archaeology, augmented where available with data on their activity on GitHub. We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations. Our results show an uneven adoption of open source collaborative practices beyond the basic use of `git` as a version control system and GitHub to host source code. A majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base. More community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories. Preliminary findings from ongoing work also shows how collaborative behaviours vary across different subcommunities. By analyzing interactions among archaeologists using the GitHub platform, including the ways they contribute to collective code bases, we demonstrate how archaeological software engineering is beginning to foster new kinds of collaborative commitments that blend open source protocols, DIY work ethic, conventional forms of scholarly communication, and cultural practices unique to archaeology.
bibliography: openarchaeo-collaboration.bib
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---

@@

library("fs")
library("geomtextpath")
library("ggiraph")
library("ggraph")
library("gt")
library("here")
library("khroma")
library("lubridate")
library("patchwork")
library("tidygraph")
library("tidyverse")
```

>>>>

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo (https://open-archaeo.info/), a curated list of 500+ active open projects in archaeology, augmented where available with data on their activity on GitHub. We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations. Our results show an uneven adoption of open source collaborative practices beyond the basic use of `git` as a version control system and GitHub to host source code. A majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base. More community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories. Preliminary findings from ongoing work also shows how collaborative behaviours vary across different subcommunities. By analyzing interactions among archaeologists using the GitHub platform, including the ways they contribute to collective code bases, we demonstrate how archaeological software engineering is beginning to foster new kinds of collaborative commitments that blend open source protocols, DIY work ethic, conventional forms of scholarly communication, and cultural practices unique to archaeology.
bibliography: references.bib
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---

@@

library("fs")
library("geomtextpath")
library("ggiraph")
library("ggraph")
library("gt")
library("here")
library("khroma")
library("lubridate")
library("patchwork")
library("scales")
library("tidygraph")
library("tidyverse")
```
